Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-08 ind_files/04_unification_discussion.tex
Minor phrasing changes

Stochastic finite population models often exhibit behaviors that are markedly different from their deterministic limits. Since real-life populations are stochastic and finite, it is thus imperative that modellers work with stochastic first-principles models instead of their deterministic limits, lest they risk missing important phenomena that are unique to stochastic systems \citep{black_stochastic_2012,schreiber_does_2022,hastings_transients_2004,shoemaker_integrating_2020}. In the context of eco-evolutionary theory, \cite{doebeli_towards_2017} have called for a reformulation of evolutionary dynamics starting from stochastic birth-death processes on the grounds that such a formulation is more fundamental and mechanistic, a view that has also been voiced by theorists before \citep{metcalf_why_2007,geritz_mathematical_2012}. In this thesis, I present and derive from first principles some fundamental relations (equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance}) that any such reformulation must satisfy. Further, these relations deal with biologically important quantities, lend themselves to simple biological interpretation, and are very general. They thus fulfill the criteria to be called `fundamental theorems' (in the sense of \cite{queller_fundamental_2017}), or `unifying principles' (in the sense of \cite{lion_theoretical_2018}) for the dynamics of finite populations. \cite{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' of the lot, and equations like the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in our framework - Equation \eqref{nD_stochastic_RM} is the most fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula (in principle for any moment of the distribution of $f$ in the population, though this quickly becomes too tedious to actually carry out in practice). Just like in infinite population models, the various terms of equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} lend themselves to simple biological interpretation. Note that actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that high dimensional evolutionary birth-death models can exhibit a dizzying array of complicated phenomena, including limit cycles and evolutionary chaos \citep{doebeli_diversity_2017}. My framework also works with unstructured populations, neglecting any population structure due to age, class, developmental stage, or space, all of which can lead to complex dynamics and greatly reduce analytical tractability. Lastly, I have also neglected any potential complications introduced by genotype-phenotype maps and genetic processes such as dominance and epistasis since they can be absorbed into the per-capita birth and death rates and the equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} will hold for arbitrarily complicated birth and death rates as long as they are of the functional form \eqref{nD_functional_forms_for_replicator}. Like the Price equation, the utility of these equations lies not in their solutions, but instead in their generality and the fact that their terms help us clearly understand the various forces operating in biological populations \citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. The general approach of working with `high-level' processes without specifying system-specific details is part of a broader pursuit of `model-independent' eco-evolutionary theory that has recently been gaining popularity in the literature  \citep{grafen_formal_2014, queller_fundamental_2017, doebeli_towards_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022}. Such model-independent descriptions that `abstract away' system-specific details almost inevitably come at the cost of precision \citep{levins_strategy_1966, potochnik_idealization_2018}, and are thus intended to complement rather than supersede specific models of specific systems.\\

>>>>

Stochastic finite population models often exhibit behaviors that are markedly different from their deterministic limits. Since real-life populations are stochastic and finite, it is thus imperative that modellers work with stochastic first-principles models instead of their deterministic limits, lest they risk missing important phenomena that are unique to stochastic systems \citep{black_stochastic_2012,schreiber_does_2022,hastings_transients_2004,shoemaker_integrating_2020}. In the context of eco-evolutionary theory, \cite{doebeli_towards_2017} have called for a reformulation of evolutionary dynamics starting from stochastic birth-death processes on the grounds that such a formulation is more fundamental and mechanistic, a view that has also been voiced by theorists before \citep{metcalf_why_2007,geritz_mathematical_2012}. In this thesis, I present and derive from first principles some fundamental relations (equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance}) that any such reformulation must satisfy. Further, these relations deal with biologically important quantities, lend themselves to simple biological interpretation, and are very general. They thus fulfill the criteria to be called `fundamental theorems' (in the sense of \cite{queller_fundamental_2017}), or `unifying principles' (in the sense of \cite{lion_theoretical_2018}) for the dynamics of finite populations. \cite{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' of the lot, and equations like the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in our framework - Equation \eqref{nD_stochastic_RM} is the most fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula (in principle for any moment of the distribution of $f$ in the population, though this quickly becomes too tedious to actually carry out in practice).If we additionally assume that the quantity $f$ follows a Gaussian distribution in the population, then the mean and variance completely characterize the distribution of $f$, and equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} together specify the complete stochastic dynamics of the system. Note that even in this case, actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that in high dimensions, evolutionary birth-death models can exhibit a dizzying array of complicated phenomena, including limit cycles and evolutionary chaos \citep{doebeli_diversity_2017}. My framework also works with unstructured populations, neglecting any population structure due to age, class, developmental stage, or space, all of which can lead to complex dynamics and greatly reduce analytical tractability. Lastly, I have also neglected any potential complications introduced by genotype-phenotype maps and genetic processes such as dominance and epistasis since they can be absorbed into the per-capita birth and death rates and our equations hold for arbitrarily complicated birth and death rates as long as they are of the functional form \eqref{nD_functional_forms_for_replicator}. However, despite these practical difficulties, just like in infinite population models, the various terms of equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} always lend themselves to simple biological interpretation. Thus, like the Price equation, the utility of these equations lies not in their solutions, but instead in their generality and the fact that their terms help us clearly understand the various forces operating in biological populations \citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. The general approach of working with `high-level' processes without specifying system-specific details is part of a broader pursuit of `model-independent' eco-evolutionary theory that has recently been gaining popularity in the literature  \citep{grafen_formal_2014, queller_fundamental_2017, doebeli_towards_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022}. Such model-independent descriptions that `abstract away' system-specific details almost inevitably come at the cost of precision \citep{levins_strategy_1966, potochnik_idealization_2018}, and are thus intended to complement rather than supersede specific models of specific systems.\\
